Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2013-08-26 paper.tex
add description of the implicit rotation update

We also modify slightly the correction done to the gravitational
source terms after integrating the system in time.  In the original
Castro paper (Eqs. 19 and 20), the state after the conservative update
due to fluxes and external sources (other than gravity) was
represented with the superscript `$(2,\star)$'.  The gravitational
sources were then corrected, making them centered in time, yielding
the final state at the end of the hydrodynamics step, indicated with
the superscript `$(2)$'.  This appeared as:
\begin{eqnarray}
(\rho {\bf u})^{(2)} &=& (\rho {\bf u})^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf g})^{(2,\star)} - 
                              (\rho {\bf g})^{(1)} \right ]\\
(\rho E)^{(2)} &=& (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2,\star)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{eqnarray}
Here we make the subtle change of using the corrected ${\bf u}$ in the
correction to the energy.  So our new energy correction equation appears as:
\begin{equation}
(\rho E)^{(2)} = (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{equation}
Both of these are second-order accurate, but we've found that the
latter is slightly better for conserving energy.  This change is now
the default in the public version of Castro (it can be controlled by
{\tt castro.grav\_source\_type}.

@@

>>>>

@@

We also modify slightly the correction done to the gravitational
source terms after integrating the system in time.  In the original
Castro paper (Eqs. 19 and 20), the state after the conservative update
due to fluxes and external sources (other than gravity) was
represented with the superscript `$(2,\star)$'.  The gravitational
sources were then corrected, making them centered in time, yielding
the final state at the end of the hydrodynamics step, indicated with
the superscript `$(2)$'.  This appeared as:
\begin{eqnarray}
(\rho {\bf U})^{(2)} &=& (\rho {\bf U})^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf g})^{(2,\star)} - 
                              (\rho {\bf g})^{(1)} \right ]\\
(\rho E)^{(2)} &=& (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf U\cdot g})^{(2,\star)} - 
                              (\rho {\bf U\cdot g})^{(1)} \right ]
\end{eqnarray}
Here we make the subtle change of using the corrected ${\bf u}$ in the
correction to the energy.  So our new energy correction equation appears as:
\begin{equation}
(\rho E)^{(2)} = (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf U\cdot g})^{(2)} - 
                              (\rho {\bf U\cdot g})^{(1)} \right ]
\end{equation}
Both of these are second-order accurate, but we've found that the
latter is slightly better for conserving energy.  This change is now
the default in the public version of Castro (it can be controlled by
{\tt castro.grav\_source\_type}.

Rotation is handled similarly to gravity during the conservative update.
The momentum equation, with rotation sources appears as:
\begin{equation}
\frac{\partial \rho {\bf U}}{\partial t} + \nabla \cdot (\rho {\bf U U} + p) = 
   \rho {\bf g} - 2 \rho {\bf \Omega \times U} 
                - \rho {\bf \Omega \times (\Omega \ times U)}
\end{equation}
we take the rotation axis to be in the $z$-direction, ${\bf \Omega} =
\Omega \hat{\bf k}$.  Working out the cross-products, the rotation
forces appear as
\begin{equation}
{\bf F}_\mathrm{rot} = \rho \left (
    \begin{array}{c}
     \phantom{-}2 \Omega v + \Omega^2 x \\
              - 2 \Omega u + \Omega^2 y \\
              0 
    \end{array}
  \right )
\end{equation}
We want to time-center this force in the final conservative update, giving
updates for the $u$ and $v$ components of the velocity as:
\begin{eqnarray}
\frac{(\rho u)^{n+1} - (\rho u)^n}{\Delta t} + A^{(x),n+1/2} &=&
   \frac{1}{2} \left [ (\rho {\bf g}\cdot \hat{\bf i})^n
                      + (\rho {\bf g}\cdot \hat{\bf i})^{n+1} \right ] \nonumber\\&+&
   \frac{1}{2} \left [ (2\rho \Omega v)^n + (2\rho \Omega v)^{n+1} \right ] \nonumber \\&+&
   \frac{1}{2} \left [ (\rho \Omega^2 x)^n + (\rho \Omega^2 x)^{n+1} \right ] \\
\frac{(\rho v)^{n+1} - (\rho v)^n}{\Delta t} + A^{(y),n+1/2} &=&
   \frac{1}{2} \left [ (\rho {\bf g}\cdot \hat{\bf j})^n +
                       (\rho {\bf g}\cdot \hat{\bf j})^{n+1} \right ] \nonumber \\ &+&
   \frac{1}{2} \left [ (-2\rho \Omega u)^n + (-2\rho \Omega u)^{n+1} \right ] \nonumber \\ &+&
   \frac{1}{2} \left [ (\rho \Omega^2 y)^n + (\rho \Omega^2 y)^{n+1} \right ]
\end{eqnarray}
where here the explicit advective term from the Godunov method is represented as $A$ and is already time-centered.  In the Castro algorithm, a full $\Delta t$ of the old-time-level source is first added and later we correct this by subtracting off $\Delta t/2$ of the old-time-level term and adding $\Delta t/2$ of the new-time-level source.  This correction appears as:
\begin{eqnarray}
(\rho u)^{n+1} = (\rho u)^\prime 
   &+& \underbrace{
      \frac{\Delta t}{2} \left [ (\rho {\bf g}\cdot \hat{\bf i})^{n+1}
                               - (\rho {\bf g}\cdot \hat{\bf i})^{n} \right ]
    }_{\equiv \Delta^{g,x}} \nonumber \\
   &+& \frac{\Delta t}{2} \left [ (2\rho \Omega v)^{n+1} 
                                - (2\rho \Omega v)^{n} \right ] \nonumber \\
   &+& \underbrace{
      \frac{\Delta t}{2} \left [ (\rho \Omega^2 x)^{n+1} 
                               - (\rho \Omega^2 x)^{n} \right ]
    }_{\equiv \Delta^{c,x}} \\
%
(\rho v)^{n+1} = (\rho v)^\prime 
   &+& \underbrace{
      \frac{\Delta t}{2} \left [ (\rho {\bf g}\cdot \hat{\bf j})^{n+1}
                               - (\rho {\bf g}\cdot \hat{\bf j})^{n} \right ]
    }_{\equiv \Delta^{g,y}} \nonumber \\
   &+& \frac{\Delta t}{2} \left [ (-2\rho \Omega u)^{n+1} 
                                - (-2\rho \Omega u)^{n} \right ] \nonumber \\
   &+& \underbrace{
      \frac{\Delta t}{2} \left [ (\rho \Omega^2 y)^{n+1} 
                               - (\rho \Omega^2 y)^{n} \right ]
    }_{\equiv \Delta^{c,y}} \\
\end{eqnarray}
For compactness, we define the following:
\begin{eqnarray}
S_x &=& (\rho u)^\prime + \Delta^{g,x} + \Delta^{c,x} 
        - \frac{\Delta t}{2} (2 \rho \Omega v)^n \\
%
S_y &=& (\rho v)^\prime + \Delta^{g,y} + \Delta^{c,y} 
        + \frac{\Delta t}{2} (2 \rho \Omega u)^n 
\end{eqnarray}
and our updates are now a coupled implicit system for the updates of
$u$ and $v$:
\begin{eqnarray}
(\rho u)^{n+1} = S_x + \frac{\Delta t}{2}(2\rho \Omega v)^{n+1} \label{eq:rhouupdate}\\
(\rho v)^{n+1} = S_y - \frac{\Delta t}{2}(2\rho \Omega u)^{n+1} 
\end{eqnarray}
This can be solved analytically:
\begin{equation}
(\rho v)^{n+1} = \frac{S_y - \Delta t \Omega S_x}{1 + \Delta t^2 \Omega^2}
\end{equation}
and then we can find $(\rho u)^{n+1}$ from Eq.~\ref{eq:rhouupdate}.
\clearpage
